---
title: "Assignment"
author: "Mary"
date: "2025-06-26"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "Adult Income Dataset: From Exploration to Statistical Analysis"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
library(broom)
library(modelsummary)
``` 

## 1. Load and Clean Data

```{r load-data}
adult <- read_csv("adult.csv", show_col_types = FALSE) |>
  clean_names() |>
  mutate(across(where(is.character), as.factor))
```

## 2. Basic Exploration

```{r glimpse-skim}
glimpse(adult)
skim(adult)
```

### Missing Values

```{r missing-values}
adult |> summarise(across(everything(), ~ sum(is.na(.))))
```

### Class Balance

```{r income-balance}
adult |> count(income) |> mutate(prop = n / sum(n))
```

```{r}
datasummary_skim(adult %>% select(age, gender, hours_per_week, education),
                 title = "Descriptive Statistics: Adult Income Variables")
```


## 3. Visual Exploration

### Hours per Week by Income
```{r box-hours-income}
ggplot(adult, aes(x = income, y = hours_per_week, fill = income)) +
  geom_boxplot() +
  theme_minimal()install.packages("modelsummary")
```

### Age Histogram by Sex
```{r age-hist-sex}
ggplot(adult, aes(x = age, fill = gender)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  theme_minimal()
```

### Education vs Hours (Colored by Income)
```{r edu-vs-hours}
ggplot(adult, aes(x = education, fill = income)) +
  geom_bar(position = "dodge") +
  labs(title = "Income Distribution by Education Level",
       x = "Education Level", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 4. Descriptive Statistics

```{r summary-stats}
adult |> 
  group_by(income) |>
  summarise(
    count = n(),
    mean_age = mean(age),
    median_hours = median(hours_per_week),
    sd_hours = sd(hours_per_week)
  )
```

## 5. Hypothesis Testing

### T-Test: Hours per Week by Income
```{r ttest-hours}
t.test(hours_per_week ~ income, data = adult)
```

### T-Test: Age by Sex
```{r ttest-age-sex}
t.test(age ~ gender, data = adult)
```

### Chi-Square: Education vs Income
```{r chisq-edu-income}
table_edu <- table(adult$education, adult$income)
chisq.test(table_edu)
```

## 6. ANOVA: Age ~ Education
```{r anova-age-edu}
aov_model <- aov(age ~ education, data = adult)
summary(aov_model)
```

## 7. Logistic Regression

```{r logistic-model}
model <- glm(income ~ age + education + hours_per_week + gender, 
             data = adult, family = binomial)
summary(model)
```

### Odds Ratios
```{r model-odds}
tidy(model, exponentiate = TRUE, conf.int = TRUE)
```

### Accuracy
```{r model-accuracy}
predicted <- ifelse(predict(model, type = "response") > 0.5, 
                    ">50K", "<=50K")
mean(predicted == adult$income)
```

## 8. Conclusion

- Individuals earning >50K work significantly more hours per week.
- Age, education level, hours per week, and sex are significant predictors of high income.
- The logistic regression model has moderate accuracy and provides interpretable odds ratios.
